---
title: "Linear interpolation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Interpolation function: definition

From the R documentation
```{r, eval=FALSE, echo=TRUE}
help(approx)
```
**approx(x, y=NULL, xout, method="linear", n=50, rule = 1)**  

**x, y**	vectors giving the coordinates of the points to be interpolated  

**xout**	an optional set of values specifying where interpolation is to take place  

**n**	If xout is not specified, interpolation takes place at n equally spaced points spanning the interval [min(x), max(x)]  

**rule**	an integer describing how interpolation is to take place outside the interval [min(x), max(x)]. If rule is 1 then NAs are returned for such points and if it is 2, the value at the closest data extreme is used

## Using "n"

```{r, eval=TRUE, echo=TRUE}
a <- seq(-4,5,by=1)

f <- function(x){(x^2)-x}

parabola <- f(a)

par(mfrow=c(2,2))

for(i in c(10,20,30,40)){
  plot(parabola,xaxt='n',pch=19,col="black",xlab="",main=paste0("n=",i))
  axis(1,at=c(1:length(a)),labels=a)
  points(approx(parabola,n=i),pch=22,col="red")
}
```

## Using "xout"

```{r, eval=TRUE, echo=TRUE}
abscissa <- c(-1,length(a)+2)
abscissa_points <- seq(-1,length(a)+2,by=1)
abscissa_labels <- seq(min(a)-2,max(a)+2,by=1)
ordinate <- c(min(parabola),max(parabola)+3)

plot(parabola,pch=19,col="black",xaxt="n", xlim=abscissa, ylim=ordinate, xlab="",main="xout=abscissa_points")
axis(1, at=abscissa_points, labels=abscissa_labels)
points(approx(parabola,xout=abscissa_points),pch=22,col="red")

par(mfrow=c(1,2))

begin <- abscissa_points[1]-1
finish <- abscissa_points[length(abscissa_points)]+1

for(set_rule in c(1,2)){
  plot(parabola,pch=19,col="black",xaxt="n", xlim=abscissa, ylim=ordinate, xlab="",main=paste0("outside the x range, rule=",set_rule))
  axis(1, at=abscissa_points, labels=abscissa_labels)
  points(approx(parabola,xout=seq(begin,finish,by=1),rule=set_rule),pch=22,col="red")
}
```
