---
title: "Principal Component Analysis"
author: "Fabio Natalini"
date: "2023-01-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressWarnings(suppressPackageStartupMessages(library(knitr)))
suppressWarnings(suppressPackageStartupMessages(library(kableExtra)))
```

## Random data set

First, let’s generate a data set with continuous variables. We will make *ad-hoc* variables to make our example straightforward. 

```{r}
n <- 300
lapply(c(9,18,27), function(x){
  set.seed(1)
  cbind(
    rnorm(n,mean=x-1,sd=1),
    rnorm(n,mean=x,sd=1),
    rnorm(n,mean=x+1,sd=1)
  ) |> as.data.frame() |> setNames(nm=paste0("_",x,c("minus","norm","plus")))
}) |> do.call(what=cbind) -> data.set
```

```{r}
data.set %>% kable(row.names = TRUE) %>%
  kable_styling("basic",full_width = FALSE) %>%
  scroll_box(height = "200px")
```
<br>
```{r}
par(xpd=TRUE,mar=c(1,1,1,8),cex=0.8)
matplot(data.set,type="l",col=rainbow(ncol(data.set)),ylab="",xlab="",xaxt="n",yaxt="n")
legend("topright", inset=c(-0.22, 0),lty=1,legend=names(data.set),col=rainbow(ncol(data.set)))
```

The variables are clearly grouped. This should lead to clear results.

## PCA

We will use the function `princomp` to perform a principal component analysis. Then we will examine the proportion of variance explained by each component, the scree plot of the eigenvalues and the component matrix graph.

```{r}
data.pca <- princomp(data.set)
summary(data.pca) #Proportions of Variance
plot(data.pca,type='l',main='scree plot') #scree plot of eigenvalues
biplot(data.pca) #plot the component matrix graph
```

The results shows that there are 3 important components. Let’s take them and plot the loadings against the original variables.

```{r}
main.comp <- data.pca$loadings[,1:3]
# data.pca$scores[,1:3] #these are the scores of the first components 

par(mfrow=c(3,1),cex.axis=0.9,cex.main=0.9,las=2,mar=c(1,4,2,1))
barplot(unname(main.comp[,1]),main="Comp.1",ylab="loadings")
barplot(unname(main.comp[,2]),main="Comp.2",ylab="loadings")
par(mar=c(5,4,2,1))
barplot(main.comp[,3],main="Comp.3",ylab="loadings")
```

The 3 selected components indicate the underlying grouping we expected.
